Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2014-11-11 paper/main.tex
done with RC of abstract and intro

\title{Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}}
% Certified Resource Bounds for C
% Compositional Certified Resource Bounds for C
% Compositional Quantitative Resource Analysis of C Programs
% Compositional Resource-Bound Certification for C Programs

@@

\begin{abstract}
  This paper presents a new approach for automatically deriving
  worst-case resource bounds for C programs.
  %
  The described technique combines ideas from amortized analysis and
  abstract interpretation in a unified framework to address four
  challenges for state-of-the-art techniques: compositionality, user
  interaction, generation of proof certificates, and scalability.
  %
  \emph{Compositionality} is achieved by incorporating the potential
  method of amortized analysis.  It enables the derivation of global
  whole-program bounds with local derivation rules by naturally
  tracking size changes of variables in sequenced loops and function
  calls.  The resource consumption of functions is described abstractly
  and a function call can be analyzed without access to the function
  body.
  %\
  \emph{User interaction} is supported with a new mechanism that
  clearly separates qualitative and quantitative verification. A user
  can guide the analysis to derive complex non-linear bounds by using
  auxiliary variables and assertions.  The assertions are
  separately proved using established qualitative techniques such as
  abstract interpretation or Hoare logic.
  %
  \emph{Proof certificates} are automatically generated from the local
  derivation rules. A soundness proof of the derivation system with
  respect to a formal cost semantics guarantees the validity of the
  certificates.
  %
  \emph{Scalability} is attained by an efficient reduction of
  bound inference to a linear optimization problem that can be solved
  by off-the-shelf LP solvers.
  %
  The analysis framework is implemented in the publically-available
  tool $\toolname$. An experimental evaluation demonstrates the
  advantages of the new technique with a comparison of $\toolname$ with
  existing tools on challenging micro benchmarks and the analysis of
  more than XXX lines of C code from the XXX benchmark suite.

@@

Available techniques for deriving worst-case resource bounds
fall into two categories.  Techniques
in the first category derive impressive bounds for
numerical imperative programs, but are not fully compositional.
This is problematic if one needs to derive global whole-program bounds.
Techniques in the second category derive tight whole-program
bounds on functional programs that operate on inductive data structures.
They are highly compositional, scale for large
programs, and work directly on the syntax.  However, % despite all
% these theoretical and practical advantages,
it has been
a long-time open problem to make them work in the
presence of signed integers, mutation, and non-linear
control flow. As a result, they are not suitable for
typical imperative code.

Notable tools in the first category include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14}.
They lack compositionality in at least three ways.
First, they all base their analysis on some form of
\emph{ranking function}.
Since this
is usually a local analysis and loop bounds are
arithmetic expressions that depend on the values of
variables just before the loop.
This makes it hard to give a
resource bound on a sequence of loops and procedure calls in
terms of the input parameters of a procedure.
Second, while all popular imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract resource behavior
and have to inline the procedure body to perform their
analysis.  Third, when a tool fails to find a resource
bound for a program, there is no possibility for sound
user interaction with the tool other than to the rewrite
the program. For example, there is no concept of manual proofs of
resource bounds and no framework that could support composition
of manually derived bounds with automatically inferred bounds.
One reason for this lack of interaction is that available tools
work on an abstract version of the source program. This
abstracted program has sometimes more behaviors than the
original one (introduced by non-determinism) and is usually
expressed in a language that does not feature memory operations.

Tools in the second category originate form the potential method of
amortized analysis and type systems for functional
programs~\cite{Jost03,HoffmannAH12}. It has been shown that class
definitions of object oriented programs~\cite{Jost06} and
data-structure predicates of separation logic~\cite{Atkey10} can play
the role of the type system in imperative programs. However, these
techniques fail for numerical imperative programs because these programs lack
fine-grained data-structure information to hook the analysis on.
Another issue is that existing techniques associate \emph{potential}
directly with single program variables.  In imperative programs, this
fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
$y - i$ decreases, but not $|i|$.  Mutation is yet another problem.
While functional programs relate multiple fixed inputs (the typing
context) with one output (the expression typed), imperative programs
can change the value of variables in scope, so the potential of these
variables has to be updated after mutations to ensure soundness.

This paper presents a new framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two previous
categories: It solves the compositionality issues of techniques
for numerical imperative code by adapting amortized-analysis based
techniques from the functional world.
%
Our automated analysis is able to infer linear
resource bounds on C programs with mutually-recursive functions and integer loops.
%
The resource behavior of functions can be summarized
in a function specification that can be used at every
call-side without accessing the function body.
%
To our knowledge this is the first technique based on
amortized analysis that is able to derive bounds that depend on negative numbers
and differences of variables.  It is also the first
resource analysis technique for C that deals naturally with
recursive functions and sequenced loops, and can handle resources that can
become available during execution (e.g., when freeing
memory).  Compared to more classical
apporaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:

\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}

The main innovation that makes amortized analysis work
on imperative languages is to base the analysis on a
Hoare-like logic and to track multivariate quantities instead
of program variables.  This leads to precise bounds
expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
intervals.
%
A distinctive feature of our analysis system is that
it reduces bound derivation to a linear optimization problem
that can solved by off-the-shelf LP solvers.  This enables the
efficient inference of global bounds for larger programs.

The use of the potential method of amortized analysis makes user
interaction possible in different ways. For one thing, we can directly
combine the new automatic analysis with manually derived bounds in a
previously-developed quantitative Hoare logic~\cite{veristack14} (see
Section XXX).  For another thing, we describe a new mechanism that
allows the separation of quantitative and qualitative verification
(see Section XXX).  Using this mechanism, the user can guide the
analysis by using auxiliary variables and logical assertions that can
be verified by existing qualitative tools such as Hoare logic or
abstract interpretation.  In this way, we can benefit from existing
automation techniques and provide a middle-ground between fully
automatic and fully manual verification for bound derivation. This
enables the semi-automatic inference of non-linear bounds, such as
polynomial, logarithmic, and exponential bounds.

We have implemented the automatic analysis in the tool
\toolname and experimentally evaluated its effectiveness by
analyzing system code and examples from the literature.
\toolname has automatically derived global resource bounds for
more than XXX lines of C code from the XXX benchmark suite.
%
\iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 1 of these patterns, and in
most cases the derived constant factors are tight.  To compare our
automatic analyzer with existing techniques, we tested our examples
with tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.
%
Several micro benchmarks demonstrate the practicality and
expressiveness of the user guided bound inference.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds a maximal element in an array, and a linear
bound that amortizes the cost of $k$ successive increments to a binary
counter.

% Give a description of the sections here.                    XXX
% Jan: We don't need that. Usually, people don't read it.

In summary, we make the following \emph{contributions}.

\begin{itemize}
\item We develop the first automatic amortized analysis for C
  programs. It is naturally compositional, tracks size changes of
  variables to derive global bounds, can handle mutually-recursive
  functions, generates resource specifications for functions, derives
  easily verifiable certificates, and handles resources that can
  become available during an execution.
\item We show how to reduce the inference of \emph{linear} resource
    bounds to efficient LP solving.
\item We describe a new method of harnessing existing qualitative
  verification techniques to guide the automatic amortized analysis
  to derive \emph{non-linear} resource bounds with LP solving.
\item We prove the soundness of the analysis with respect to a
  parametric cost semantics for C programs. The cost model can be
  further customized with function calls (\code{tick(n)}) that
  indicate resource usage. % This mechanism is also covered by the
  % soundness proof.
\item We present experiments with a prototype implementation on more
  than XXX lines of C code. A detailed comparison shows that our
  prototype is the only tool that can derive global bounds for larger
  C programs while being as powerful as existing tools when deriving
  linear local bounds for tricky loop and recursion patterns.
\end{itemize}

>>>>

\title{Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}}
% Certified Resource Bounds for C
% Compositional Certified Resource Bounds for C
% Compositional Quantitative Resource Analysis of C Programs
% Compositional Resource-Bound Certification for C Programs

@@

\begin{abstract}
  This paper presents a new approach for automatically deriving
  worst-case resource bounds for C programs.
  %
  The described technique combines ideas from amortized analysis and
  abstract interpretation in a unified framework to address four
  challenges for state-of-the-art techniques: compositionality, user
  interaction, generation of proof certificates, and scalability.
  %
  \emph{Compositionality} is achieved by incorporating the potential
  method of amortized analysis.  It enables the derivation of global
  whole-program bounds with local derivation rules by naturally
  tracking size changes of variables in sequenced loops and function
  calls.  The resource consumption of functions is described abstractly
  and a function call can be analyzed without access to the function
  body.
  %\
  \emph{User interaction} is supported with a new mechanism that
  clearly separates qualitative and quantitative verification. A user
  can guide the analysis to derive complex non-linear bounds by using
  auxiliary variables and assertions.  The assertions are
  separately proved using established qualitative techniques such as
  abstract interpretation or Hoare logic.
  %
  \emph{Proof certificates} are automatically generated from the local
  derivation rules. A soundness proof of the derivation system with
  respect to a formal cost semantics guarantees the validity of the
  certificates.
  %
  \emph{Scalability} is attained by an efficient reduction of
  bound inference to a linear optimization problem that can be solved
  by off-the-shelf LP solvers.
  %
  The analysis framework is implemented in the publically-available
  tool $\toolname$. An experimental evaluation demonstrates the
  advantages of the new technique with a comparison of $\toolname$ with
  existing tools on challenging micro benchmarks and the analysis of
  more than 2000 lines of C code from the cBench benchmark suite.

@@

Available techniques for automatically deriving worst-case resource bounds
fall into two categories.  Techniques
in the first category derive impressive bounds for
numerical imperative programs, but are not fully compositional.
This is problematic if one needs to derive global whole-program bounds.
Techniques in the second category derive tight whole-program
bounds for programs with regular loop and recursion patterns that
decrease the size of an individual variable or data structure.
They are highly compositional, scale for large
programs, and work directly on the syntax.  However, % despite all
% these theoretical and practical advantages,
it has been
a long-time open problem to make them work for typical imperative code
with non-regular iteration patterns, signed integers, mutation, and non-linear
control flow. % As a result, they are not suitable for
% typical imperative code.

Tools in the first category include
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14}.
They lack compositionality in at least two ways.
First, they all base their analysis on some form of
\emph{ranking function} or \emph{counter instrumentation}
that is linked to a local analysis.  As a result, loop bounds are
arithmetic expressions that depend on the values of
variables just before the loop.
This makes it hard to give a
resource bound on a sequence of loops and function calls in
terms of the input parameters of a function.
Second, while all popular imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract resource behavior
and have to inline the procedure body to perform their
analysis.

Tools in the second category originate form the potential method of
amortized analysis and type systems for functional
programs~\cite{Jost03,HoffmannAH12}. It has been shown that class
definitions of object oriented programs~\cite{Jost06} and
data-structure predicates of separation logic~\cite{Atkey10} can play
the role of the type system in imperative programs. However, these
techniques fail for numerical imperative programs because these programs lack
fine-grained data-structure information to hook the analysis on.
Another issue is that existing techniques associate \emph{potential}
directly with single program variables.  In imperative programs, this
fails for loops as simple as \lstinline{for(i=x;i<y;i++)} where
$y - i$ decreases, but not $|i|$. %  Mutation is yet another problem.
% While functional programs relate multiple fixed inputs (the typing
% context) with one output (the expression typed), imperative programs
% can change the value of variables in scope, so the potential of these
% variables has to be updated after mutations to ensure soundness.

A general problem with existing tools is user interaction. When a tool
fails to find a resource bound for a program, there is no possibility
for sound user interaction to guide the tool during during bound
derivation.
For example, there is no concept of manual proofs of
resource bounds and no framework that could support composition
of manually derived bounds with automatically inferred bounds.

% Finally, the tools often operate on an abstract version of the and use
% complex external tools from term rewriting and abstract
% interpretation. This makes it hard to generate a proof certificate for
% the bounds or even to prove the soundness of the tools with respect to
% a formal cost semantics.

This paper presents a new framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two previous
categories: It solves the compositionality issues of techniques
for numerical imperative code by adapting amortized-analysis based
techniques from the functional world.
%
Our automated analysis is able to infer
resource bounds on C programs with mutually-recursive functions and integer loops.
%
The resource behavior of functions can be summarized
in a function specification that can be used at every
call-side without accessing the function body.
%
To our knowledge this is the first technique based on
amortized analysis that is able to derive bounds that depend on negative numbers
and differences of variables.  It is also the first
resource analysis technique for C that deals naturally with
recursive functions and sequenced loops, and can handle resources that can
become available during execution (e.g., when freeing
memory).  Compared to more classical
apporaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  Using only one
simple mechanism, it handles:
\itemskip
\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
\itemskip
The main innovation that makes amortized analysis work
on imperative languages is to base the analysis on a
Hoare-like logic and to track multivariate quantities instead
of program variables.  This leads to precise bounds
expressed as functions of sizes $|[x, y]| = \max(0, y-x)$ of
intervals.
%
A distinctive feature of our analysis system is that
it reduces linear bound inference  to a linear optimization problem
that can solved by off-the-shelf LP solvers.  This enables the
efficient inference of global bounds for larger programs.
%
Moreover, our local inference rules automatically generate
proof certificates that can be easily checked in linear time.

The use of the potential method of amortized analysis makes user
interaction possible in different ways. For one thing, we can directly
combine the new automatic analysis with manually derived bounds in a
previously-developed quantitative Hoare logic~\cite{veristack14} (see
\pref{sec:soundness}).  For another thing, we describe a new mechanism that
allows the separation of quantitative and qualitative verification
(see \pref{sec:anno}).  Using this mechanism, the user can guide the
analysis by using auxiliary variables and logical assertions that can
be verified by existing qualitative tools such as Hoare logic or
abstract interpretation.  In this way, we can benefit from existing
automation techniques and provide a middle-ground between fully
automatic and fully manual verification for bound derivation. This
enables the semi-automatic inference of non-linear bounds, such as
polynomial, logarithmic, and exponential bounds.

We have implemented the analysis system in the tool
\toolname and experimentally evaluated its effectiveness by
analyzing system code and examples from the literature.
\toolname has automatically derived global resource bounds for
more than 2000 lines of C code from the cBench benchmark suite.
%
\iffull{\pref{app:cat}}{The extended version
  of this article~\cite{anon_extended}}{} contains more than $30$
challenging loop and recursion patterns that we collected from open
source software and the literature.  Our analysis can find
asymptotically tight bounds for all but 1 of these patterns, and in
most cases the derived constant factors are tight.  To compare $\toolname$ 
with existing techniques, we tested our examples
with tools such as KoAT~\cite{BrockschmidtEFFG14},
Rank~\cite{AliasDFG10}, and LOOPUS~\cite{SinnZV14}.  Our experiments
show that the bounds that we derive are often more precise than those
derived by existing tools.  Only LOOPUS~\cite{SinnZV14}, which also
uses amortization techniques, is able to achieve a similar precision.

Several micro benchmarks demonstrate the practicality and
expressiveness of the user guided bound inference.  For example, we
derive a logarithmic bound for a binary search function, a bound that
depends on the contents of an array to describe the exact cost of a
function that finds a maximal element in an array, and a bound 
that amortizes the cost of $k$ successive increments to a binary
counter (see \pref{sec:anno}).

In summary, we make the following \emph{contributions}.
\itemskip
\begin{itemize}
\item We develop the first automatic amortized analysis for C
  programs. It is naturally compositional, tracks size changes of
  variables to derive global bounds, can handle mutually-recursive
  functions, generates resource abstractions for functions, derives
  proof certificates, and handles resources that can
  become available during execution.
\item We show how to automatically reduce the inference of \emph{linear} resource
    bounds to efficient LP solving.
\item We describe a new method of harnessing existing qualitative
  verification techniques to guide the automatic amortized analysis
  to derive \emph{non-linear} resource bounds with LP solving.
\item We prove the soundness of the analysis with respect to a
  parametric cost semantics for C programs. The cost model can be
  further customized with function calls (\code{tick(n)}) that
  indicate resource usage. % This mechanism is also covered by the
  % soundness proof.
\item We present experiments with a prototype implementation on more
  than 2000 lines of C code. A detailed comparison shows that our
  prototype is the only tool that can derive global bounds for larger
  C programs while being as powerful as existing tools when deriving
  linear local bounds for tricky loop and recursion patterns.
\end{itemize}
\itemskip
%
{\footnotesize \sl
[To the reviewers: The auxiliary material contains an anonymized
version of the extended article, and the source code of $\toolname$
and all examples.] }
